---
title: "From Deprivation to Degree"
subtitle: "How Income Affects Youth Education in English Towns"
author: "Group R you OK? ：Vicky Feng · Ruyu Li · Dev Karan Gur"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    toc: true
    incremental: true
    transition: slide
    center: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
---

```{r package loading, include=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(knitr)
library(kableExtra)
```

## Introduction

**Regional inequality in education** is a persistent concern in England.

This study examines **whether town-level income deprivation influences academic attainment for youth aged 18 to 22**.

We focus on three key outcome stages:

-   **Level 3 qualification at age 18**
-   **University entry at age 19**
-   **Level 6+ qualification by age 22**

## Research Question

**Does the income level of a town affect the educational outcomes of young people?**

## Variables and Dataset Description

```{r data loading}
edu_df <- read_csv("../data/english_education.csv")
```

```{r related variables chosen}
selected_vars <- c( 
  "income_flag",
  "level_3_at_age_18",
  "activity_at_age_19_full_time_higher_education",
  "highest_level_qualification_achieved_by_age_22_level_6_or_above",
  "size_flag"
)

edu_df_selected <- edu_df %>%
  select(all_of(selected_vars))%>%
  rename(
    level3_at_18 = level_3_at_age_18,
    uni_at_19 = activity_at_age_19_full_time_higher_education,
    level6_at_22 = highest_level_qualification_achieved_by_age_22_level_6_or_above
  )
```

Based on `english_education.csv` from the TidyTuesday project and **UK Office for National Statistics (ONS)**.

Contains **1104 town-level observations** and **31 variables**.

**Key related variables** used in this analysis :

```{r}
#| label: tbl-variable-description
#| tbl-cap: "Key Variables in the Analysis"

edu_variable_desc <- data.frame(
  Variable = c("income_flag", "level3_at_18", "uni_at_19", "level6_at_22", "size_flag"),
  Description = c(
    "Income deprivation group (higher / mid / lower)",
    "% achieving Level 3 qualification at age 18",
    "% entering full-time university at age 19",
    "% attaining Level 6+ qualification by age 22",
    "Town size classification (small / medium / large)"
  )
)

kable(edu_variable_desc, format = "markdown", align = "l")
```

## Table of Mean and SD of Educational Outcomes by Income Group

```{r data cleaning}
edu_df_clean <- edu_df_selected %>%
  filter(!is.na(income_flag) & income_flag != "Cities") %>%
  mutate(income_flag = factor(income_flag,
                              levels = c("Higher deprivation towns",
                                         "Mid deprivation towns",
                                         "Lower deprivation towns")))
```

```{r tabel of mean and SD}
#| label: tbl-edu_means_by_income
#| tbl-cap: "Mean and SD of Educational Outcomes by Income Group(Cities Excluded)"

table1 <- edu_df_clean %>%
  group_by(income_flag) %>%
  summarise(
    mean_level3 = mean(level3_at_18, na.rm = TRUE),
    sd_level3 = sd(level3_at_18, na.rm = TRUE),
    mean_uni19 = mean(uni_at_19, na.rm = TRUE),
    sd_uni19 = sd(uni_at_19, na.rm = TRUE),
    mean_level6 = mean(level6_at_22, na.rm = TRUE),
    sd_level6 = sd(level6_at_22, na.rm = TRUE)
  ) %>%
  rename(
    "Income Group" = income_flag,
    "Mean: Level 3 at 18" = mean_level3,
    "SD: Level 3 at 18" = sd_level3,
    "Mean: HE at 19" = mean_uni19,
    "SD: HE at 19" = sd_uni19,
    "Mean: Level 6 at 22" = mean_level6,
    "SD: Level 6 at 22" = sd_level6
  )

kable(table1, digits = 2) %>%
  kable_styling(full_width = FALSE)

```

**Mean values improve** as deprivation decreases

**Standard deviations** are **smaller** in lower deprivation towns → more consistent outcomes

## Educational Outcomes by Income Group (Boxplot)

```{r educational outcomes boxplot}
#| label: fig-edu_outcomes_boxplot
#| fig-cap: "Educational Outcomes by Income Group(Cities Excluded)"

edu_long <- edu_df_clean %>%
  pivot_longer(
    cols = c(level3_at_18,
             uni_at_19,
             level6_at_22),
    names_to = "education_stage",
    values_to = "value"
  )

edu_long$income_flag <- recode(edu_long$income_flag,
  "Higher deprivation towns" = "Higher\ndeprivation\ntowns",
  "Mid deprivation towns" = "Mid\ndeprivation\ntowns",
  "Lower deprivation towns" = "Lower\ndeprivation\ntowns"
)

edu_long$education_stage <- recode(edu_long$education_stage,
  "level3_at_18" = "Level 3 at 18",
  "uni_at_19" = "HE at 19",
  "level6_at_22" = "Level 6 at 22"
)

ggplot(edu_long, aes(x = income_flag, y = value, fill = income_flag)) +
  geom_boxplot() +
  facet_wrap(~education_stage, scales = "free_y") +
  labs(x = "Income Group", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 9, lineheight = 1.1))
```

**Lower** deprivation towns have: **Higher medians** and **Narrower interquartile ranges**

**Higher** deprivation towns show: **Greater variability** and **Lower educational outcomes**

**Clear trend**: Income linked to educational success

## Results

```{r educational attainment barplot}
#| label: fig-income-education-by-size
#| fig-cap: "Educational Attainment by Income Deprivation Level and Town Size (Cities Excluded)"
#| fig-subcap:
#| - "Comparison of educational outcomes across income levels and town sizes"

summary_df <- edu_df_clean %>%
  group_by(income_flag, size_flag) %>%
  summarise(
    level3 = mean(level3_at_18, na.rm = TRUE),
    uni19 = mean(uni_at_19, na.rm = TRUE),
    level6 = mean(level6_at_22, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(level3, uni19, level6),
    names_to = "education_stage",
    values_to = "mean_value"
  )

summary_df$education_stage <- recode(summary_df$education_stage,
  "level3" = "Level 3 at Age 18",
  "uni19" = "University at 19",
  "level6" = "Level 6+ by Age 22"
)

summary_df$income_flag <- recode(summary_df$income_flag,
  "Higher deprivation towns" = "Higher\ndeprivation\ntowns",
  "Mid deprivation towns" = "Mid\ndeprivation\ntowns",
  "Lower deprivation towns" = "Lower\ndeprivation\ntowns"
)

ggplot(summary_df, aes(x = income_flag, y = mean_value, fill = education_stage)) +
  geom_col(position = "dodge", width = 0.7) +
  facet_wrap(~size_flag) +
  labs(
    x = "Income Deprivation Group",
    y = "Average Proportion",
    fill = "Education Stage"
  ) +
  scale_fill_manual(
    values = c(
      "Level 3 at Age 18" = "#4C72B0",
      "University at 19" = "#55A868",
      "Level 6+ by Age 22" = "#C44E52"
    )
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 9, lineheight = 1.1))
```

Educational attainment **rises** as income deprivation decreases.

**Lower deprivation towns outperform** others at ages 18, 19, and 22.

The gap is most pronounced in **small towns**, especially at Level 6+.

Highlights the role of local economic conditions.

## Conclusions and Recommendations

Strong link: less deprived towns show better educational outcomes.

Support should target high-deprivation towns — especially small ones.

Recommend expanding dataset and integrating qualitative factors.

## Ending Page

**Thank you!**

Presentation done!

Questions welcome.
